---
title: "Spatial and Socioeconomic Determinants of Sepsis Hospitalisation Rates in France"
author: "Susy Echeverria Londono"
date: today
format:
  pdf:
    toc: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(broom)
library(sf)
library(readr)
library(here)
library(ggplot2)
library(patchwork)
library(knitr)
library(kableExtra)
library(stargazer)
library(INLA) # Bayesian spatial models (BYM2)
library(regclass)
library(spatialreg)
library(glmnet)

theme_set(theme_minimal(base_size = 12))

knitr::opts_chunk$set(
  dev = "ragg_png",
  dpi = 120,
  fig.width = 7,
  fig.height = 5,
  out.width = "88%",
  fig.retina = 1
)
```

# Introduction

This analysis examined the geographic distribution of sepsis hospitalisation rates across French regions and investigated whether regional socioeconomic conditions — median living standard, poverty intensity, population density, and educational attainment — could explain the observed variation. A sequential analytical strategy was applied, moving from spatial exploratory analysis through frequentist regression to Bayesian hierarchical modelling, with each step designed to address a specific alternative explanation for the results of the previous one.

For this analysis, I am focusing on metropolitan France + Corse only (n = 13 regions) to avoid the confounding influence of overseas territories, which have very different healthcare contexts and sepsis epidemiology.

# Spatial clustering
## Global spatial clustering (Moran's I)

```{r}
#| include: false
# echo = FALSE

moran_mc <- readRDS("data/derived/06_moran_mc_result.RDS")
```


```{r}
# echo = FALSE

print(moran_mc)
```

- Global Moran's I was positive (I = `{r} round(moran_mc$statistic[[1]],2)`), suggesting a weak tendency toward spatial clustering of sepsis hospitalisation rates, but this did not reach statistical significance (p = `{r} round(moran_mc$p.value, 2)`), likely reflecting the limited power of the test at the regional scale (n = 13). 

## Local spatial clustering (LISA)

However, given the extraordinary population density and urban concentration of Ile-de-France, I am interested in exploring if Ile-de-France behaves as a driver of local clusters of low incidence of sepsis (healthcare spill-over) or as an outlier. To explore this, I am performing Local Moran's I (LISA) analysis. 

```{r}
#| echo: false

LISA_results <- readRDS("data/derived/06_LISA_results.RDS")

LISA_results %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "center")

```

- Île-de-France appears in the LH quadrant of the Moran scatterplot — suggesting lower-than-average sepsis rates surrounded by higher-than-average neighbors — though this local pattern did not reach statistical significance (p > 0.05), likely reflecting the limited power of the test at the regional scale.

- Pays de la Loire is the only statistically significant cluster (p = 0.034) — a genuine LL cluster, meaning low sepsis rates in a region surrounded by other low-rate regions (Bretagne, Normandie, Centre-Val de Loire). 

- It is important to note that this spatial autocorrelation test (Moran's I) with n = 13 regions has low statistical power, therefore, the results of this analysis does NOT prove the absence of spatial structure — it may simply reflect insufficient data to detect moderate autocorrelation.


# Links between sepsis hospitalisation rates and INSEE territorial indicators. 

To examine the relationship between regional sepsis hospitalisation rates and socioeconomic predictors, I applied a sequential modelling approach of increasing complexity. I first used Ordinary Least Squares (OLS) regression as a baseline

## Ordinary Least Squares (OLS) regression


```{r}
#| echo: false

ols_model <- readRDS("data/derived/06_ols_model.RDS")

stargazer(ols_model, type="text", title="Results", align=TRUE)

```

```{r}
#| echo: false

coef_df <- 
  tidy(ols_model, conf.int = TRUE, conf.level = 0.95) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = recode(term,
      niveau_vie_z             = "Median living standard",
      intensite_pauvrete_pct_z = "Poverty intensity",
      pct_non_scol_z           = "Low educational attainment",
      log_dens_z               = "Population density (log)"
    ),
    term      = reorder(term, estimate),
    Sig       = ifelse(p.value < 0.05, "p < 0.05", "p ≥ 0.05")
  )

p1 <- ggplot(coef_df, aes(x = estimate, y = term, colour = Sig)) +
  geom_vline(xintercept = 0,
             linetype = "dashed", colour = "grey60", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_colour_manual(
    values = c("p < 0.05" = "firebrick", "p ≥ 0.05" = "steelblue")
  ) +
  labs(
    title    = "OLS regression coefficients",
    subtitle = "Error bars are 95% confidence intervals",
    x        = "Coefficient (hospitalisation rate per 100k, per 1 SD)",
    y        = NULL,
    colour   = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey50", size = 10),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.y      = element_text(size = 11)
  )

```

- OLS regression revealed no significant association between regional socioeconomic indicators and sepsis hospitalisation rates (F(4,8) = 0.96, p = 0.48; adjusted R² = −0.01). 

- Severe collinearity between population density and low educational attainment (r = −0.91) inflated standard errors for individual coefficients. I therefore use the Ridge regression to assess whether coefficient stabilisation under penalisation would recover any meaningful signal.

## Ridge regression

```{r}
#| echo: false

ridge_model <- readRDS("data/derived/06_ridge_model.RDS")
```

```{r}
#| echo: false
coef(ridge_model, s = "lambda.min")

best_lambda <- ridge_model$lambda.min
cat("Best lambda:", best_lambda)
```

```{r}
#| echo: false

coef_summary <- readRDS("./06_ridge_coeff_boots.RDS")

n_boot = 200

ggplot(coef_summary,
             aes(x = Estimate, y = Predictor)) +
  geom_vline(xintercept = 0,
             linetype = "dashed", colour = "grey60", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper),
                 height = 0.2, colour = "steelblue", linewidth = 0.8) +
  geom_point(size = 3, colour = "steelblue") +
  labs(
    title    = "Ridge regression coefficients at optimal λ",
    subtitle = paste0(
      "λ = ", round(ridge_model$lambda.min, 1),
      " — error bars are bootstrap 95% intervals (", n_boot, " resamples)"
    ),
    x = "Coefficient",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey50", size = 10),
    panel.grid.minor = element_blank(),
    axis.text.y      = element_text(size = 11)
  )
```


- Ridge regression with cross-validated penalty ($\lambda$ = 2630) shrunk all coefficients to near zero, confirming that the weak OLS signal was not an artefact of multicollinearity. 

- Even after stabilising the coefficient estimates, socioeconomic indicators collectively explained little variation in regional sepsis hospitalisation rates. 

- The direction of economic predictors (income negative, poverty positive) is consistent with expectation, but the magnitudes are too small to support effects conclusions. 

- Population density and low educational achievement showed no association with sepsis rates even after their collinearity was addressed.


## Bayesian hierarchical model (BYM2)

Given the small number of regions (n = 13) and the count-derived nature of the outcome -- number of hospitalisations, I performed a Bayesian hierarchical model (BYM2) as a complementary analysis using R-INLA. This approach provides posterior risk estimates for each region with full uncertainty quantification, and accommodates the Poisson data-generating process more appropriately than OLS. 

```{r}
#| echo: false

BYM2_model_res <- readRDS("./data/derived/07_BYM2_model_results.RDS")

kable(
  BYM2_model_res,
  format    = "pipe",
  col.names = c("Predictor", "Mean (log RR)", "95% Credible Interval", "Rate Ratio"),
  caption   = "BYM2 model — fixed effects on regional sepsis hospitalisation rates"
)
```


```{r}
#| echo: false
bym2_model <- readRDS("data/derived/07_BYM2_model.RDS")

kable(
  bym2_model$summary.hyperpar,
  format    = "pipe", 
  caption   = "BYM2 model — Posterior summaries of BYM2 model hyperparameters"
)
```

```{r}
#| echo: false
# Extract marginals for each fixed effect
fe_names <- rownames(bym2_model$summary.fixed)
fe_names <- fe_names[fe_names != "(Intercept)"]

fe_marginals <- lapply(fe_names, function(nm) {
  marg <- bym2_model$marginals.fixed[[nm]]
  df   <- as.data.frame(marg)
  names(df) <- c("x", "density")
  df$Predictor <- nm
  df
})

fe_df <- bind_rows(fe_marginals) %>%
  mutate(
    Predictor = recode(Predictor,
                       niveau_vie_z             = "Median living standard",
                       intensite_pauvrete_pct_z = "Poverty intensity",
                       pct_non_scol_z           = "Low educational attainment",
                       log_dens_z               = "Population density (log)"
    ),
    # Flag area under curve to left or right of zero
    side = ifelse(x < 0, "negative", "positive")
  )

p4 <- ggplot(fe_df, aes(x = x, y = density)) +
  geom_area(aes(fill = side), alpha = 0.4) +
  geom_line(linewidth = 0.7, colour = "grey30") +
  geom_vline(xintercept = 0,
             linetype = "dashed", colour = "grey40", linewidth = 0.5) +
  scale_fill_manual(
    values = c("negative" = "#2166AC", "positive" = "#D6604D"),
    guide  = "none"
  ) +
  facet_wrap(~ Predictor, scales = "free", ncol = 2) +
  labs(
    title    = "Posterior distributions of fixed effects — BYM2 model",
    subtitle = "Blue = negative effect region; red = positive effect region.\nDashed line = null (log RR = 0)",
    x        = "log Rate Ratio",
    y        = "Posterior density"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey50", size = 10),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold", size = 10)
  )

p4
```

- Posterior credible intervals for all fixed effects spanned zero, consistent with OLS and Ridge results. i.e., non significant association between the sepsis rate and socioeconomical predictors. 

- The high precision of the spatial random effects (posterior mean = 41.5) indicated minimal residual regional heterogeneity after covariate adjustment. However, this modest variance detects regional differences on the rate ratio scale.

- The spatial mixing parameter φ = 0.47 (95% CI: 0.04–0.95) was unidentifiable given the sample size, which shows any reliable partition of variance between spatially structured and unstructured components. i.e., we cannot conclude if geographical patterns are driven by spatial proximity or region-specific factors. 


```{r}
#| echo: false
map_df <- readRDS("./data/derived/07_bym2_spatial.RDS")
re_df <- readRDS("./data/derived/07_bym2_re.RDS")

p2 <- ggplot(map_df) +
  geom_sf(aes(fill = RR), colour = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low      = "#2166AC",   # blue = lower than average
    mid      = "#F7F7F7",   # white = average
    high     = "#D6604D",   # red = higher than average
    midpoint = 1,
    name     = "Posterior\nRR",
    limits   = c(
      floor(min(map_df$RR, na.rm = TRUE) * 10) / 10,
      ceiling(max(map_df$RR, na.rm = TRUE) * 10) / 10
    )
  ) +
  labs(
    title    = "Posterior relative risk of sepsis hospitalisation",
    subtitle = "BYM2 model — metropolitan France. RR > 1 (red) = above average; RR < 1 (blue) = below average",
    caption  = "Overseas territories excluded. Spatial weights: queen contiguity."
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey50", size = 10),
    plot.caption  = element_text(colour = "grey60", size = 8),
    legend.position = "right"
  )

p2 

```


```{r}
#| echo: false
re_plot_df <- re_df %>%
  mutate(
    region_name = reorder(region_name, RR),
    sig = case_when(
      RR_lower > 1 ~ "Above average",
      RR_upper < 1 ~ "Below average",
      TRUE         ~ "Inconclusive"
    )
  )

p3 <- ggplot(re_plot_df,
             aes(x = RR, y = region_name, colour = sig)) +
  geom_vline(xintercept = 1,
             linetype = "dashed", colour = "grey50", linewidth = 0.6) +
  geom_errorbarh(aes(xmin = RR_lower, xmax = RR_upper),
                 height = 0.3, linewidth = 0.7) +
  geom_point(size = 3) +
  scale_colour_manual(
    values = c(
      "Above average" = "#D6604D",
      "Below average" = "#2166AC",
      "Inconclusive"  = "grey50"
    )
  ) +
  labs(
    title    = "Posterior relative risk by region — BYM2 model",
    subtitle = "Points = posterior mean RR; bars = 95% credible intervals.\nDashed line = national average (RR = 1)",
    x        = "Relative Risk",
    y        = NULL,
    colour   = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey50", size = 10),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.y      = element_text(size = 10)
  )

p3 
```

- Even though minimal residual regional heterogeneity was found in the BYM2 model, Bourgogne-Franche-Comté and Grand Est showed posterior RRs credibly above the national average, reflecting raw sepsis rates sufficiently extreme to persist after Bayesian shrinkage toward the mean. Centre-Val de Loire is highlighted as the region with posterior RRs credibly below the national average. 


# Model comparison

```{r}
#| echo: false

comparison_table2<- read.csv("./outputs/tables/model_comparison_results.csv")

kable(comparison_table2 %>% filter(Method != "LM Diagnostics"), format = "pipe",
      caption   = "Summary of conclusions — spatial and regression analyses of regional sepsis hospitalisation rates in metropolitan France (n = 13 regions)")
```

## Main conclusions

- Across all methods, the results converged on the same finding: no socioeconomic predictor showed a statistically or probabilistically credible association with regional sepsis hospitalisation rates.

- These results should not be interpreted as evidence that socioeconomic conditions are irrelevant to sepsis outcomes in France. The administrative regional scale is too coarse to detect the mechanisms through which poverty, education, and urbanisation influence sepsis rates. 

- With only 13 observations, all models operated at the boundary of statistical identifiability, and even moderate true effects would have been undetectable given the available power. 

- Bourgogne-Franche-Comté and Grand Est showed posterior RRs credibly above the national average. These regions show regional risk that deviate enough from the mean to be detectable even at this sample size. These findings reinforce the call for finer-scale analysis at the département level to understand the mechanisms underlying elevated risk in eastern France. 

- Finer socioeconomic granularity would provide the statistical power necessary to detect and characterise the geographic drivers of sepsis hospitalisation in France. The observed variation in sepsis rates could reflect processes operating at a finer geographic scale, perhaps case-mix differences between sepsis subtypes, and commune or departemental level socioeconomic heterogeneity that is obscured by regional aggregation.

# ANNEX:
```{r}
#| eval: false
#| include: false

dat_sf <- readRDS("./data/derived/06_spatial_all_dat.RDS")
ols_fit <- readRDS("./data/derived/06_ols_model.RDS")

# Test residuals for spatial autocorrelation
dat_sf$resid <- residuals(ols_fit)

## mapping residuals to check for spatial patterns in the residuals (which would suggest that the OLS model is missing spatially structured predictors or that a spatial model is needed)

ggplot(dat_sf) +
  geom_sf(aes(fill = resid)) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0,
    name = "Residual"
  ) +
  labs(title = "OLS Residuals — Sepsis hospitalisation rate",
       subtitle = "Red = higher than predicted, Blue = lower than predicted") +
  theme_minimal()
```


```{r}
#| echo: false
# OLS vs Ridge coefficient comparison ----
# directly shows shrinkage effect

# Extract OLS coefficients
ols_coefs <- tidy(ols_model, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  select(term, estimate, conf.low, conf.high) %>%
  mutate(Model = "OLS")

# Extract Ridge coefficients at optimal lambda
ridge_est <- as.data.frame(
  as.matrix(coef(ridge_model, s = "lambda.min"))
) %>%
  rename(estimate = 1) %>%
  filter(rownames(.) != "(Intercept)") %>%
  mutate(
    term      = rownames(.),
    conf.low  = NA_real_,   # Ridge has no standard CIs
    conf.high = NA_real_,
    Model     = "Ridge"
  )

combined_coefs <- bind_rows(ols_coefs, ridge_est) %>%
  mutate(
    term = recode(term,
      niveau_vie_z             = "Median living standard",
      intensite_pauvrete_pct_z = "Poverty intensity",
      pct_non_scol_z           = "Low educational attainment",
      log_dens_z               = "Population density (log)"
    ),
    term = reorder(term, estimate)
  )

p4 <- ggplot(combined_coefs,
             aes(x = estimate, y = term, colour = Model, shape = Model)) +
  geom_vline(xintercept = 0,
             linetype = "dashed", colour = "grey60", linewidth = 0.6) +
  geom_errorbarh(
    data = filter(combined_coefs, Model == "OLS"),
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.15, linewidth = 0.8
  ) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("OLS" = "steelblue", "Ridge" = "firebrick")) +
  scale_shape_manual(values  = c("OLS" = 16, "Ridge" = 17)) +
  labs(
    title    = "OLS vs Ridge regression coefficients",
    subtitle = paste0(
      "OLS: 95% confidence intervals shown. ",
      "Ridge: λ = ", round(ridge_model$lambda.min, 1),
      ", no intervals (shrinkage estimator)"
    ),
    x      = "Coefficient",
    y      = NULL,
    colour = NULL,
    shape  = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey50", size = 10),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.y      = element_text(size = 11)
  )

p4

```

