---
title: "Regional analysis of sepsis hospitalisations in France"
author: "Susy Echeverria Londono"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(sf)
library(readr)
library(here)
library(ggplot2)
library(patchwork)

theme_set(theme_minimal(base_size = 12))
```

```{r}
#| include: false
source("./R/fun.R")
regions_joined <- readRDS("./data/derived/01_sepsis_spatial.RDS")
```


## Spatial description of sepsis

### 1. Nombre de séjours pour sepsis des habitants de la région pour 100 000 habitants

```{r}
#| echo: false
#| fig-cap: Nombre de séjours pour sepsis des habitants de la région pour 100 000 habitants
#| fig-height: 6
#| fig-width: 10

plot_maps_var(regions_joined, 
              var = "tx_sejours_sepsis_100k", 
              label_var = "Nombre de séjours pour sepsis des habitants de la région pour 100 000 habitants", 
              shared_legend = FALSE)

```

Variable	Interpretation
Sepsis sans choc	Incidence of diagnosed sepsis
Septic shock	Severe progression / hospital severity
Combined	Total burden of sepsis hospitalizations

Reasoning:
A region with high sepsis but low shock, suggests earlier detection or better management. Region that shows lower incidence but higher shock, suggests delayed diagnosis or different healthcare access.


```{r}
#| echo: false
#| fig-cap: Nombre de séjours pour sepsis des habitants de la région pour 100 000 habitants
#| fig-height: 6
#| fig-width: 10


## summ over the types of sepsis to get total sepsis incidence
all_sepsis <- 
  regions_joined %>%
  group_by(NOM_M) %>%
  summarise(tx_sejours_sepsis_100k = sum(tx_sejours_sepsis_100k, na.rm = TRUE))

plot_fr_var_insets(all_sepsis, 
                   var = "tx_sejours_sepsis_100k", 
                   label = "Nombre de séjours pour sepsis des habitants de la région pour 100 000 habitants", 
                   group_level = NULL)

```


### 2. pourcentage de décès

Important nuance

Because the data are based on hospital stays rather than unique patients, the indicator reflects:

mortality per hospitalization episode, not mortality per patient. = l’âge moyen des patients hospitalisés pour sepsis dans cette région, calculé à partir de l’ensemble des séjours hospitaliers enregistrés.

Le pourcentage de décès (pct_deces) correspond à la proportion de séjours hospitaliers pour sepsis qui se terminent par un décès. Il s’agit donc d’un indicateur de mortalité hospitalière parmi les hospitalisations pour sepsis.

```{r}
#| echo: false
#| fig-cap: Mortalité hospitalière des séjours pour sepsis (%)
#| fig-height: 6
#| fig-width: 10

plot_maps_var(regions_joined, 
              var = "pct_deces", 
              label_var = "Mortalité hospitalière des séjours pour sepsis (%)", 
              shared_legend = FALSE)

```

### 3. Durée moyenne de séjour (DMS)

Durée moyenne de séjour (DMS) is useful to interpret regional differences in sepsis burden and severity.

So DMS helps you understand whether a high number of sepsis stays reflects: many relatively short admissions, or fewer but more severe and resource-intensive admissions.

Sepsis with longer average stays may indicate: more severe infections, complications, ICU care, slower recovery.

So if a region has both: high sepsis rate, and long DMS this may suggest a heavier clinical burden, not just more admissions.

DMS complements incidence by showing intensity of care use.

```{r}
#| echo: false
#| fig-cap: Durée moyenne de séjour (DMS)
#| fig-height: 6
#| fig-width: 10

plot_maps_var(regions_joined, 
              var = "dms_nuitees", 
              label_var = "Durée moyenne de séjour (DMS)",
              shared_legend = FALSE)

```

```{r}
#| echo: false
#| fig-cap: âge moyen des patients hospitalisés
#| fig-height: 6
#| fig-width: 10

plot_maps_var(regions_joined, 
              var = "age_moyen", 
              label_var = "âge moyen des patients hospitalisés",
              shared_legend = FALSE)

```


## Association between sepsis incidence and INSEE variable

The exercise was implemented using three sepsis outcomes, analysed separately and in aggregate as requested. Because the final question specifically concerns homogeneous regional profiles based on septic shock rates and territorial indicators, clustering and regional typology were centered on the septic shock rate, while analyses of non-shock and total sepsis were used as complementary descriptive and interpretive comparisons

### Spatial description of INSEE variables
```{r}
#| echo: false

insee_spatial <- readRDS("./data/derived/05_insee_spatial.RDS")

plot_fr_var_insets(insee_spatial, 
                   var = "pop_non_scol15", 
                   label = "population non scolarisée", 
                   group_level = NULL)

```

```{r}
#| echo: false

plot_fr_var_insets(insee_spatial, 
                   var = "niveau_vie_median_annuel", 
                   label = "Poverty", 
                   group_level = NULL)
```

### Association plots
```{r}
#| echo: false
```


### Final regional profiles / clustering (septic shock)
```{r}
#| echo: false
```

### Final regional profiles / clustering (total sepsis, optionally non-shock)
```{r}
#| echo: false
```
