---
title: "Spatial analysis of septic shock hospitalisation rates across French regions"
author: "Susy Echeverria Londono"
date: today
format:
  pdf:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(sf)
library(readr)
library(here)
library(ggplot2)
library(patchwork)
library(knitr)
library(kableExtra)

theme_set(theme_minimal(base_size = 12))

knitr::opts_chunk$set(
  dev = "ragg_png",
  dpi = 120,
  fig.width = 7,
  fig.height = 5,
  out.width = "88%",
  fig.retina = 1
)
```

# Introduction

To better characterise the links between regional sepsis incidence and the selected INSEE territorial indicators, I am performing two complementary analytical approaches in this report: 

1. Besag-York-Mollié (BYM) Bayesian spatial model: 


# Objectives

Stage 1 — Spatial autocorrelation diagnosis (Moran's I) + OLS + SAR/SEM on metropolitan France only. 
Stage 2 — Regularized regression (Ridge/LASSO) to understand which INSEE indicators matter most.



# caveats:  

- spatial autocorrelation test (Moran's I): with n = 13 regions, this test has low statistical power. A non-significant result does NOT prove the absence of spatial structure — it may simply reflect insufficient data to detect moderate autocorrelation.


# Spatial autocorrelation diagnosis (Moran's I)

```{r}
# echo = FALSE

moran_mc <- readRDS("data/derived/06_moran_mc_result.RDS")

plot(moran_mc, main = "Moran's I permutation distribution (septic shock rate)")
```


Global Moran's I was positive (I = `{r} round(moran_mc$statistic[[1]],2)`), suggesting a weak tendency toward spatial clustering of sepsis hospitalisation rates, but this did not reach statistical significance (p = `{r} round(moran_mc$p.value, 2)`), likely reflecting the limited power of the test at the regional scale (n = 13). These results neither confirm nor rule out spatial autocorrelation.

However, given the extraordinary population density and urban concentration of Ile-de-France, I am interested in exploring if Ile-de-France behaves as a driver of local clusters of low incidence of sepsis (healthcare spill-over) or as an outlier. To explore this, I am performing Local Moran's I (LISA) analysis. 

```{r}
#| echo: false

LISA_ile_de_france <- readRDS("data/derived/06_LISA_ile_de_france.RDS")

print(LISA_ile_de_france)

```


Île-de-France appears in the LH quadrant of the Moran scatterplot — suggesting lower-than-average sepsis rates surrounded by higher-than-average neighbors — though this local pattern did not reach statistical significance (p > 0.05), likely reflecting the limited power of the test at the regional scale.

This result presents Île-de-France as a notable region: despite having the highest population density in France, it shows a below-average sepsis hospitalisation rate, while its neighboring regions tend to show higher rates. This pattern is consistent with the hypothesis that concentrated healthcare infrastructure in Île-de-France may contribute to lower sepsis incidence locally, though it does not appear to extend to surrounding regions. 


