---
title: "Regional analysis of sepsis hospitalisations in France"
author: "Susy Echeverria Londono"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(sf)
library(readr)
library(here)
library(ggplot2)
library(patchwork)

theme_set(theme_minimal(base_size = 12))
```


```{r}
#| include: false
source("./R/fun.R")
regions_joined <- readRDS("./data/derived/01_sepsis_spatial.RDS")
```

## Sepsis incidence
### 1. Nombre de séjours pour sepsis des habitants de la région pour 100 000 habitants
```{r}
#| echo: false
#| fig-cap: Nombre de séjours pour sepsis des habitants de la région pour 100 000 habitants
#| fig-height: 6
#| fig-width: 10

plot_maps_var(regions_joined, 
              var = "tx_sejours_sepsis_100k", 
              label_var = "Nombre de séjours pour sepsis des habitants de la région pour 100 000 habitants", 
              shared_legend = FALSE)

```

### 2. pourcentage de décès 
Important nuance

Because the data are based on hospital stays rather than unique patients, the indicator reflects:

mortality per hospitalization episode, not mortality per patient. = l’âge moyen des patients hospitalisés pour sepsis dans cette région, calculé à partir de l’ensemble des séjours hospitaliers enregistrés.

Le pourcentage de décès (pct_deces) correspond à la proportion de séjours hospitaliers pour sepsis qui se terminent par un décès. Il s’agit donc d’un indicateur de mortalité hospitalière parmi les hospitalisations pour sepsis.


```{r}
#| echo: false
#| fig-cap: Mortalité hospitalière des séjours pour sepsis (%)
#| fig-height: 6
#| fig-width: 10

plot_maps_var(regions_joined, 
              var = "pct_deces", 
              label_var = "Mortalité hospitalière des séjours pour sepsis (%)", 
              shared_legend = FALSE)

```

### 3. Durée moyenne de séjour (DMS)

Durée moyenne de séjour (DMS) is useful to interpret regional differences in sepsis burden and severity.

So DMS helps you understand whether a high number of sepsis stays reflects: many relatively short admissions, or fewer but more severe and resource-intensive admissions.

Sepsis with longer average stays may indicate: more severe infections, complications, ICU care, 
slower recovery.

So if a region has both: high sepsis rate, and long DMS this may suggest a heavier clinical burden, not just more admissions.

DMS complements incidence by showing intensity of care use.

```{r}
#| echo: false
#| fig-cap: Durée moyenne de séjour (DMS)
#| fig-height: 6
#| fig-width: 10

plot_maps_var(regions_joined, 
              var = "dms_nuitees", 
              label_var = "Durée moyenne de séjour (DMS)",
              shared_legend = FALSE)

```


```{r}
#| echo: false
#| fig-cap: âge moyen des patients hospitalisés
#| fig-height: 6
#| fig-width: 10

plot_maps_var(regions_joined, 
              var = "age_moyen", 
              label_var = "âge moyen des patients hospitalisés",
              shared_legend = FALSE)

```

## INSEE variables
## 1.  Niveau de vie et pauvreté par région




