---
title: "Spatial and Socioeconomic Determinants of Sepsis Hospitalisation Rates in France"
author: "Susy Echeverria Londono"
date: today
format:
  pdf:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(sf)
library(readr)
library(here)
library(ggplot2)
library(patchwork)
library(knitr)
library(kableExtra)
library(stargazer)

theme_set(theme_minimal(base_size = 12))

knitr::opts_chunk$set(
  dev = "ragg_png",
  dpi = 120,
  fig.width = 7,
  fig.height = 5,
  out.width = "88%",
  fig.retina = 1
)
```

# Introduction

This analysis examined the geographic distribution of sepsis hospitalisation rates across French regions and investigated whether regional socioeconomic conditions — median living standard, poverty intensity, population density, and educational attainment — could explain the observed variation. A sequential analytical strategy was applied, moving from spatial exploratory analysis through frequentist regression to Bayesian hierarchical modelling, with each step designed to address a specific alternative explanation for the results of the previous one.

For this analysis, I am focusing on metropolitan France + Corse only (n = 13 regions) to avoid the confounding influence of overseas territories, which have very different healthcare contexts and sepsis epidemiology.


# Spatial autocorrelation diagnosis (Moran's I)

```{r}
# echo = FALSE

moran_mc <- readRDS("data/derived/06_moran_mc_result.RDS")

plot(moran_mc, main = "Moran's I permutation distribution (septic shock rate)")
```


Global Moran's I was positive (I = `{r} round(moran_mc$statistic[[1]],2)`), suggesting a weak tendency toward spatial clustering of sepsis hospitalisation rates, but this did not reach statistical significance (p = `{r} round(moran_mc$p.value, 2)`), likely reflecting the limited power of the test at the regional scale (n = 13). These results neither confirm nor rule out spatial autocorrelation.

However, given the extraordinary population density and urban concentration of Ile-de-France, I am interested in exploring if Ile-de-France behaves as a driver of local clusters of low incidence of sepsis (healthcare spill-over) or as an outlier. To explore this, I am performing Local Moran's I (LISA) analysis. 

```{r}
#| echo: false

LISA_results <- readRDS("data/derived/06_LISA_results.RDS")

LISA_results %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "center")

```


Île-de-France appears in the LH quadrant of the Moran scatterplot — suggesting lower-than-average sepsis rates surrounded by higher-than-average neighbors — though this local pattern did not reach statistical significance (p > 0.05), likely reflecting the limited power of the test at the regional scale.

Pays de la Loire is the only statistically significant cluster (p = 0.034) — a genuine LL cluster, meaning low sepsis rates in a region surrounded by other low-rate regions (Bretagne, Normandie, Centre-Val de Loire). This is a stable, locally coherent low-risk pocket in western France, which may reflect shared regional characteristics such as healthcare access, population health, or socioeconomic factors that warrant further investigation.

Spatial autocorrelation test (Moran's I): with n = 13 regions, this test has low statistical power. A non-significant result does NOT prove the absence of spatial structure — it may simply reflect insufficient data to detect moderate autocorrelation.


# Links between sepsis hospitalisation rates and INSEE territorial indicators. 

To examine the relationship between regional sepsis hospitalisation rates and socioeconomic indicators from INSEE, I applied a sequential modelling approach of increasing complexity. I first used Ordinary Least Squares (OLS) regression as a baseline

## Ordinary Least Squares (OLS) regression


```{r}
#| echo: false

ols_model <- readRDS("data/derived/06_ols_model.RDS")
stargazer(ols_model, type="text", title="Results", align=TRUE)

```


OLS regression revealed no significant association between regional socioeconomic indicators and sepsis hospitalisation rates (F(4,8) = 0.96, p = 0.48; adjusted R² = −0.01). Severe collinearity between population density and low educational attainment (r = −0.91) inflated standard errors for individual coefficients. I therefore use the Ridge regression to assess whether coefficient stabilisation under penalisation would recover any meaningful signal.

## Ridge regression

```{r}
#| echo: false

ridge_model <- readRDS("data/derived/06_ridge_model.RDS")

coef(ridge_cv, s = "lambda.min")
```

Ridge regression with cross-validated penalty (λ = 2630) shrunk all coefficients to near zero, confirming that the weak OLS signal was not an artefact of multicollinearity. Even after stabilising the coefficient estimates, socioeconomic indicators collectively explained negligible variation in regional sepsis hospitalisation rates. The direction of economic predictors (income negative, poverty positive) was consistent with expectation, but the magnitudes were too small to support substantive conclusions. Population density and low educational attainment showed no independent association with sepsis rates even after their collinearity was addressed.


## Bayesian hierarchical model (BYM2)

Given the small number of regions (n = 13) and the count-derived nature of the outcome -- number of hospitalisations, I performed a Bayesian hierarchical model (BYM2) as a complementary analysis using R-INLA. This approach provides posterior risk estimates for each region with full uncertainty quantification, and accommodates the Poisson data-generating process more appropriately than OLS. 

```{r}
#| echo: false

BYM2_model_res <- readRDS("./data/derived/07_BYM2_model_results.RDS")

kable(
  BYM2_model_res,
  format    = "pipe",
  col.names = c("Predictor", "Mean (log RR)", "95% Credible Interval", "Rate Ratio"),
  caption   = "BYM2 model — fixed effects on regional sepsis hospitalisation rates"
)
```


Posterior credible intervals for all fixed effects spanned zero, consistent with OLS and Ridge results. 

```{r}
bym2_model <- readRDS("data/derived/07_BYM2_model.RDS")

print(bym2_main$summary.hyperpar)
```

- The high precision of the spatial random effects (posterior mean = 41.5) indicated minimal residual regional heterogeneity after covariate adjustment. 

- The spatial mixing parameter φ = 0.47 (95% CI: 0.04–0.95) was unidentifiable given the sample size, which shows any reliable partition of variance between spatially structured and unstructured components.


# Model comparison
```{r}
#| echo: false

comparison_table <- readRDS("./data/derived/08_model_comparison_table.RDS")
# For Word / Quarto output replace kable() call with:
kable(comparison_table, format = "pipe",
      col.names = c("Method", "Purpose", "Key Result"),
      caption   = "Summary of analytical steps — spatial and regression analyses of regional sepsis hospitalisation rates in metropolitan France (n = 13 regions)")
```
## Conclusions

1. Global Moran's I: Weak, inconclusive spatial clustering; limited power at n = 13.
2. LISA: One localised low-rate cluster in western France (Pays de la Loire: LL cluster); no widespread spatial structure. 
3. OLS Regression: No socioeconomic predictor significantly associated with sepsis rates
4. LM Diagnostics: No residual spatial dependence; SEM/SLM not warranted. 
5. Ridge Regression: Multicollinearity is not an explanation for weak fit of socioeconomic predictors. 
6. BYM2 Model: Confirms null result under correct distributional assumptions; minimal residual regional heterogeneity. 

- Across all methods, the results converged on the same finding: no socioeconomic predictor showed a statistically or probabilistically credible association with regional sepsis hospitalisation rates.

- These results should not be interpreted as evidence that socioeconomic conditions are irrelevant to sepsis outcomes in France. The administrative regional scale is too coarse to detect the mechanisms through which poverty, education, and urbanisation influence sepsis rates. 

- With only 13 observations, all models operated at the boundary of statistical identifiability, and even moderate true effects would have been undetectable given the available power. 

- The observed variation in sepsis rates could reflect processes operating at a finer geographic scale, perhaps case-mix differences between sepsis subtypes, and commune or departemental level socioeconomic heterogeneity that is obscured by regional aggregation.

- Finer socioeconomic granularity would provide the statistical power necessary to detect and characterise the geographic drivers of sepsis hospitalisation in France.



# ANEX: spatial Residual diagnostics for OLS model
```{r}
#| echo: false

# Test residuals for spatial autocorrelation
model_dat$resid <- residuals(ols_fit)

# Attach geometry back
shock_combined_resid <- 
  dat_sf %>%
  distinct(region_name, .keep_all = TRUE) %>%
  left_join(model_dat %>% select(region_name, resid), by = "region_name")


## mapping residuals to check for spatial patterns in the residuals (which would suggest that the OLS model is missing spatially structured predictors or that a spatial model is needed)

ggplot(shock_combined_resid) +
  geom_sf(aes(fill = resid)) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0,
    name = "Residual"
  ) +
  labs(title = "OLS Residuals — Sepsis hospitalisation rate",
       subtitle = "Red = higher than predicted, Blue = lower than predicted") +
  theme_minimal()
```